% abstract:
\begin{abstract}
  % context:
  Conventional data-description languages, such as context-free
  grammars, naturally define formats in which the semantic
  interpretation of a large segment of input depends hierarchically on
  the semantic interpretations of its sub-segments.
  %
  They cannot be applied to many practical and security-critical
  formats---including the Portable Document Format (PDF)---in which
  the interpretation of a segment as a \emph{Document Object Model
    (DOM)} graph depends on a concept of reference and complex
  contextual data that binds data objects to references
  \textcolor{blue}{shouldn't this be "references to data objects"?}.
  %
  Such referential context itself must often be defined
  discontinuously and compressed, to satisfy practical constraints on
  usability and performance.
  \textcolor{blue}{Sorry - I don't understand this previous sentence. How about "The integrity of these references and their context ensures that an ambiguous DOM graph is established from a basis of trust."?}

  % result:
  This paper describes a case study of a critical instance of such a
  design, namely the construction of PDF \emph{cross-reference tables},
  in the presence of potentially multiple incremental updates and multiple complex dialects expressing these references.
  %
  Over the course of our case study, we found that the full definition of
  cross-reference data in PDF contains several subtleties that are
  mis-implemented by natural implementations, but which can
  nevertheless be formalized using parser combinators in a
  data-definition language with constructs for explicitly capturing
  and updating input streams.
  %
  Our definition has served as a foundation for implementing format
  security analyses that arise naturally while constructing DOM's,
  including the analysis of document \emph{cavities} that may store
  unexamined data used to construct file \emph{polyglots}.
  
  \todo{need to be consistent in terminology: PDF Trust Chain vs Chain of Trust. Also be good to use this phrase in the Abstract. Do we quote it?}
\end{abstract}
